Exploring mathematical objects from custom-tailored mathematical universes

### 2020-06-30 paper-filmat.tex
Other aspects of toposes (in response to review)

Viewed through such a lens, a given mathematical object can have different
properties than when considered normally. In particular, it can have
better properties for the purposes of specific applications, especially if
the topos is custom-tailored to the object in question. This change of
perspective has been used in mathematical practice and demonstrates that
toposes go much beyond being logicians' testbeds. To give just a taste of what
is possible, through the lens provided by an appropriate topos, any given ring
can look like a field and hence mathematical techniques for fields also apply,
through the lens, to rings.

@@

\bigskip
\paragraph{Acknowledgments} We are grateful to Andrej Bauer, Martin
Brandenburg, Sina Hazratpour, Matthias Hutzler, Marc Nieper-Wißkirchen and
Alexander Oldenziel for invaluable discussions shaping this work, to Moritz
Laudahn, Matthias Hutzler and Milan Zerbin for their careful criticism of
earlier drafts and to Todd Lehman for preparing parts of
Figure~\ref{fig:landscape}. We thank the organizers and all participants of the
Mussomeli conference of the Italian Network for the Philosophy in Mathematics,
where this work was presented, for creating an exceptionally beautiful meeting. In
particular, we are grateful to Neil Barton, Danielle Macbeth, Gianluigi Oliveri
and Lorenzo Rossi for valuable comments.

@@

\begin{enumerate}
\item The \emph{trivial topos}. In the trivial topos, any statement whatsoever
is true. The trivial topos is not interesting on its own, but its existence
streamlines the theory and it can be an interesting question whether a given
topos coincides with the trivial topos.
\item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
in the ordinary mathematical sense.
\item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
\item $\Set^W$, the category of functors~$(W,\leq) \to \Set$ associated to
any Kripke model~$(W,\leq)$. A statement is true in this topos iff it is valid
with respect to the ordinary Kripke semantics of~$(W,\leq)$. This example shows
that the Kripke--Joyal semantics of toposes generalizes the more familiar
Kripke semantics.
\item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
if it has a \emph{computable witness} as detailed in
Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
computable, any function~$\RR \to \RR$ is continuous and the countable axiom of
choice holds (even if it does not on the meta level).
\item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
Section~\ref{sect:sheaf-toposes}. For most choices of~$X$, the
axiom of choice and the intermediate value theorem fail in~$\Sh(X)$, and this
failure is for geometric reasons.
\item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
Section~\ref{sect:smooth}. This topos contains a mirror
image of~$A$ which is a field, even if~$A$ is not.
\item $\mathrm{Bohr}(A)$, the \emph{Bohr topos} associated to a noncommutative
C\textsuperscript{*}\kern-.1ex-algebra~$A$. This topos contains a mirror image
of~$A$ which is commutative. In this sense, quantum mechanical systems (which are
described by noncommutative C\textsuperscript{*}\kern-.1ex-algebras) can be
regarded as classical mechanical systems (which are described by commutative
algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
\item $\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
and~$\psi$ are geometric formulas (formulas built from equality and specified
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$ and by arbitrary
set-indexed disjunctions~$\bigvee$).} This topos contains the
\emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
of groups contains the \emph{generic group}. Arguably it is this group
which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
The generic group has exactly those
properties which are shared by any group whatsoever.\footnote{More precisely,
this is only true for those properties which can be formulated as geometric
sequents. For arbitrary properties~$\varphi$, the statements~``the generic
group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
not be equivalent. This situation is explored
in~\cite{blechschmidt:nullstellensatz}.}
% FUTURE: Write just a little more on Set[T], that geom.impl. consequences spread.
% FUTURE: Give Set[T] its own section.
\item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
intuitionistically provable. Lambek and Scott proposed that the free topos can
reconcile moderate platonism (because this topos has a certain universal
property which can be used to single it out among the plenitude of toposes),
moderate formalism (because it is constructed in a purely syntactic way) and
moderate logicism (because, as a topos, it supports an intuitionistic type
theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
\end{enumerate}

@@

FUTURE: quick overview of the several aspects of toposes (maybe at the end, as an outlook?)
FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory
FUTURE: A topos is more than its semantics
FUTURE: Check references everywhere (give good pointers)

>>>>

Viewed through such a lens, a given mathematical object can have different
properties than when considered normally. In particular, it can have
better properties for the purposes of specific applications, especially if
the topos is custom-tailored to the object in question. This change of
perspective has been used in mathematical practice.
% and demonstrates that toposes go much beyond being logicians' testbeds.
To give just a taste of what
is possible, through the lens provided by an appropriate topos, any given ring
can look like a field and hence mathematical techniques for fields also apply,
through the lens, to rings.

@@

\bigskip
\paragraph{Other aspects of toposes} This note focuses on just a single
aspect of toposes, the view of toposes as alternate mathematical universes.
This aspect is not the only one, nor did it historically come first.

Toposes were originally conceived by Grothendieck in the early 1960s for the
needs of algebraic geometry, as a general framework for constructing and
studying invariants in classical and new geometric contexts, and it is in that
subject that toposes saw their deepest applications. The proof of Fermat's Last
Theorem is probably the most prominent such application, crucially resting on
the cohomology and homotopy invariants provided by toposes.

In the seminal work introducing
toposes~\cite{artin-grothendieck-verdier:topos}, toposes are viewed as
generalized kinds of spaces. Every topological space~$X$ gives rise to a topos,
the \emph{topos of sheaves over~$X$}, and every continuous map gives rise to a
\emph{geometric morphism} between the induced sheaf toposes, but not every
topos is of this form. While the open sets of a topological space are required
to be parts of the space, the opens of toposes are not; and while for open
subsets~$U$ and~$V$ there is only a truth value as to whether~$U$ is contained
in~$V$, in a general topos there can be many distinct ways how an open is
contained in another one. This additional flexibility is required in situations
where honest open subsets are rare, such as when studying the étale cohomology
of a scheme~\cite{milne:etale}.

That toposes could also be regarded as mathematical universes was realized only
later, by Bill Lawvere and Myles Tierney at the end of the 1960s. They abstracted some of
the most important categorical properties of Grothendieck's toposes into what
is now known as the definition of an \emph{elementary topos}. Elementary
toposes are considerably more general and less tied to geometry than the
original toposes. The theory of elementary toposes has a substantially
different, logical flavor, not least because a different notion of morphism
plays an important role. To help disambiguate, there is a trend to rename
elementary toposes to \emph{logoses}, but this text still follows the standard
convention.

A further perspective on toposes emerged in the early 1970s with the discovery
that toposes can be regarded as embodiments of a certain kind of first-order
theories, the \emph{geometric theories} briefly discussed on
page~\pageref{item:classifying-topos}. The so-called \emph{classifying toposes}
link geometrical and logical aspects and are fundamental to Olivia Caramello's
bridge-building program~\cite{caramello:tst}. Geometrically, the classifying
topos of a geometric theory~$\TT$ can be regarded as the generalized space of
models of~$\TT$; this idea is due to Monique
Hakim~\cite{hakim:relative-schemes}, though she did not cast her discovery in
this language. Logically, the classifying topos of~$\TT$ can be regarded as a
particular mathematical universe containing the \emph{generic~$\TT$-model}, a
model which has exactly those properties which are shared by all models.

Yet more views on toposes are fruitfully used --
reference~\cite{johnstone:elephants} lists ten more -- but we shall not review
them here.

\bigskip
\paragraph{Acknowledgments} We are grateful to Andrej Bauer, Martin
Brandenburg, Sina Hazratpour, Matthias Hutzler, Marc Nieper-Wißkirchen and
Alexander Oldenziel for invaluable discussions shaping this work, to Moritz
Laudahn, Matthias Hutzler and Milan Zerbin for their careful criticism of
earlier drafts and to Todd Lehman for the code for parts of
Figure~\ref{fig:landscape}. This note also profited substantially from the
thorough review by the three anonymous referees, whose efforts are much
appreciated and gladly acknowledged. We thank the organizers and all participants of the
Mussomeli conference of the Italian Network for the Philosophy in Mathematics,
where this work was presented, for creating an exceptionally beautiful meeting. In
particular, we are grateful to Neil Barton, Danielle Macbeth, Gianluigi Oliveri
and Lorenzo Rossi for valuable comments.

@@

\begin{enumerate}
\item The \emph{trivial topos}. In the trivial topos, any statement whatsoever
is true. The trivial topos is not interesting on its own, but its existence
streamlines the theory and it can be an interesting question whether a given
topos coincides with the trivial topos.
\item $\Set$, the \emph{standard topos}. A statement is true in~$\Set$ iff it is true
in the ordinary mathematical sense.
\item $\Set_M$, the topos associated to any model~$(M,\in)$ of~\textsc{zf}.
\item $\Set^W$, the category of functors~$(W,\leq) \to \Set$ associated to
any Kripke model~$(W,\leq)$. A statement is true in this topos iff it is valid
with respect to the ordinary Kripke semantics of~$(W,\leq)$. This example shows
that the Kripke--Joyal semantics of toposes generalizes the more familiar
Kripke semantics.
\item $\Eff$, the \emph{\effective topos}. A statement is true in~$\Eff$ iff
if it has a \emph{computable witness} as detailed in
Section~\ref{sect:effective-topos}. In~$\Eff$, any function~$\NN \to \NN$ is
computable, any function~$\RR \to \RR$ is continuous and the countable axiom of
choice holds (even if it does not on the meta level).
\item $\Sh(X)$, the \emph{topos of sheaves} over any space~$X$. A statement is true
in~$\Sh(X)$ iff it holds \emph{locally on~$X$}, as detailed in
Section~\ref{sect:sheaf-toposes}. For most choices of~$X$, the
axiom of choice and the intermediate value theorem fail in~$\Sh(X)$, and this
failure is for geometric reasons.
\item $\Zar(A)$, the \emph{Zariski topos} of a ring~$A$ presented in
Section~\ref{sect:smooth}. This topos contains a mirror
image of~$A$ which is a field, even if~$A$ is not.
\item $\mathrm{Bohr}(A)$, the \emph{Bohr topos} associated to a noncommutative
C\textsuperscript{*}\kern-.1ex-algebra~$A$. This topos contains a mirror image
of~$A$ which is commutative. In this sense, quantum mechanical systems (which are
described by noncommutative C\textsuperscript{*}\kern-.1ex-algebras) can be
regarded as classical mechanical systems (which are described by commutative
algebras)~\cite{butterfield-hamilton-isham:bohr,heunen-landsman-spitters:aqt}.
\item\label{item:classifying-topos}$\Set[\TT]$, the \emph{classifying topos} of a geometric theory~$\TT$.\footnote{A
geometric theory is a theory in many-sorted first-order logic whose axioms can be put as \emph{geometric
sequents}, sequents of the form~$\varphi \seq{\vec x} \psi$ where~$\varphi$
and~$\psi$ are geometric formulas (formulas built from equality and specified
relation symbols by the logical
connectives~${\top}\,{\bot}\,{\wedge}\,{\vee}\,{\exists}$ and by arbitrary
set-indexed disjunctions~$\bigvee$).} This topos contains the
\emph{generic~$\TT$-model}. For instance, the classifying topos of the theory
of groups contains the \emph{generic group}. Arguably it is this group
which we implicitly refer to when we utter the phrase ``Let~$G$ be a group.''.
The generic group has exactly those
properties which are shared by any group whatsoever.\footnote{More precisely,
this is only true for those properties which can be formulated as geometric
sequents. For arbitrary properties~$\varphi$, the statements~``the generic
group has property~$\varphi$'' and~``all groups have property~$\varphi$'' need
not be equivalent. This inbalance has mathematical applications and is explored
in~\cite{blechschmidt:nullstellensatz}.}
% FUTURE: Write just a little more on Set[T], that geom.impl. consequences spread.
% FUTURE: Give Set[T] its own section.
\item $T(\mathcal{L}_0)$, the \emph{free topos}. A statement is true in the free topos iff it is
intuitionistically provable. Lambek and Scott proposed that the free topos can
reconcile moderate platonism (because this topos has a certain universal
property which can be used to single it out among the plenitude of toposes),
moderate formalism (because it is constructed in a purely syntactic way) and
moderate logicism (because, as a topos, it supports an intuitionistic type
theory)~\cite{lambek:incompatible,couture-lambek:reflections}.
\end{enumerate}

@@

FUTURE: use the phrase "very rich and very deep" (or similarly) about set theory
FUTURE: A topos is more than its semantics
FUTURE: Check references everywhere (give good pointers)
